---
title: "ShootingIncident"
author: "Student Name (Removed for assessment)"
date: "2024-01-06"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readr)
```

## Step 1: Import data
This bellow code import data from https://catalog.data.gov/dataset

```{r import_data}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

data = read_csv(url, show_col_types = FALSE)
```

## Step 2: Tidy and Transform Data
Print a summary of the data
```{r summary}
summary(data)
```

### Select interested features
Select interested features only
```{r features}
data <- data %>%
  select(c(OCCUR_DATE, OCCUR_TIME, BORO, LOCATION_DESC, STATISTICAL_MURDER_FLAG, 
           PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE,
           Latitude, Longitude))
```

### Transform data
Convert date and time to date types
```{r date}
datat <- data %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```

There are missing data in some columns, such as `PERP_AGE_GROUP` or `PERP_SEX`, `PERP_RACE`.

There are some way to handle it:

- Replace NA with a median of the total value (e.g. age median for `PERP_AGE_GROUP`)
- Adding a new type for NA value, such as "UNKNOWN" for missing value of `PERP_SEX`

## Step 3: Add Visualizations and Analysis

### Showing number of incidents by borough
```{r borough}
data %>%
  ggplot(aes(x=BORO))+
  geom_bar(stat="count", width=0.7, fill="steelblue")+
  labs(title = "Incidents by borough", y = "Incidents") +
  theme_minimal()
```

Since there's lat/long data, let's plot it in 2D map by borou to see the spacial distribution
```{r borough_lat_long, warning=FALSE}
data %>%
  ggplot(aes(x=Longitude, y=Latitude)) + 
  geom_point(aes(color=BORO))
```

Plot the data with density

```{r borough_lat_long_density, warning=FALSE}
data %>%
  ggplot(aes(x=Longitude, y=Latitude)) + 
  geom_bin2d(bins = 100) +
  scale_fill_continuous(type = "viridis") +
  theme_bw()
```
Observation:

- There are high number of incidents in center of BROOKLYN, and between MANHATAN & BRONX

### Create a new variable for hours
```{r hour}
data <- data %>%
  mutate(hour = hour(OCCUR_TIME))
```

Plot the incident by hours
```{r hour_plot}
data %>%
  ggplot(aes(x=hour))+
  geom_line(stat="count") +
  geom_point(stat="count")
```


Observation:

- The number of incident increase significantly on evening and mid-night

Plot the incidents in hours, counting by borough

```{r hour_plot_boro}
data %>%
  ggplot(aes(x=hour, col=BORO))+
  geom_line(stat="count") +
  geom_point(stat="count")
```

### Modeling data
```{r data_total_by_hours}
data_totals_by_hour <- data %>%
  count(hour)
  
summary(data_totals_by_hour)
```

From the above visualization, let try a quadratic model between the number of incident and hour.

Firstly, create a new variable hour2:
```{r hour2}
data_totals_by_hour <- data_totals_by_hour %>%
  mutate(hour2=hour^2)
```

Then create a model
``` {r model}
quadraticModel <- lm(n ~ hour + hour2, data=data_totals_by_hour)
summary(quadraticModel)
```

Let plot the model prediction
```{r model_pred}
data_totals_by_hour_pred <- data_totals_by_hour %>%
  mutate(pred = predict(quadraticModel))

data_totals_by_hour_pred %>%
  ggplot() +
  geom_point(aes(x = hour, y = n), color = "blue") +
  geom_point(aes(x = hour, y = pred), color = "red")
```

## Conclusion
- There is a relationship between the time of the day (hour), and the chance that an shooting incident happens.
- The relation ship can be represented by a quadratic model between the hour of the day and the number of the incidents

Bias:

- People tend to think day light is safer than evening or night
- Personally, I think dense area with high population might likely to have more incidents. The future improvement could be include the population of the areas into the data set.
- I didn't check gender or race into the report. One way to improve is to consider theses factor as well.

## Session info
```{r session_info}
sessionInfo()
```
